---
title: "ICE IVIVE workflow"
Date: 06/05/2019
Note: Uses ICE2.0
output:
  pdf_document: default
  html_document:
    df_print: paged
Author: ILS
---
#Description

The workflow allows the flexibility to select from three different rat and human PK models: a 1 compartment model that incorporates Monte Carlo simulation to simulate the population variance (1C), a 3 compartment model leveraging the  [EPA's httk package](https://github.com/USEPA/CompTox-ExpoCast-httk) (Solve_3comp), and a pbpk model that is tailored for compounds with additional glucuronidation(3CGlu). The workflow is to predict the daily equivalent administered dose (EAD, mg/kg/day) that would lead to steady state blood concentration equivalent to the bioactive concentration from in vitro assays and compared to the predicted lowest effective levels (LELs) of in vivo assays, which is user provided

#load libraries

```{r}
# libraries only needed for glucuronidation models ("3CGlu", "Solve_3comp")
library(plyr) # splitting, combining and applying data
library(deSolve) # Solves for initial value problems of differential equations
library(tidyr) # helps tidy data easily
library(ggplot2) # for creating elegant complex plots
library(scales) # scaling functions for visualizing
library(foreach) # for copying functions and libraries to each cluster 
library(doParallel) # for parallelization 
```

#input variables
There are several input variables needed to run the code. Some variables are model specific, as detailed.
```{r}
f0 <- "ChemicalData_rnotebook.txt" # chemicals dats from ICE, include CASRN field as identifier
f1 <- "invitroData_xc.txt" # invitro data from ICE, includes CASRN field as identifier then acc/ac50 values
species <- "human" # human or rat
modelType <- "3CGlu" # "1C", "3CGlu", "Solve_3comp"

output_file <- "outtest2.txt"
```
For the 1 compartment model, values are needed to parameterize the Monte Carlo simulation
```{r}
nsamples <- 300 # user-provided value for the mc simulations, any number between 10 - 10,000
```
For the pbpk models some addtional parameters can be modified
```{r}
route <- "oral" # oral or iv, only needed for pbpk models ("3CGlu", "Solve_3comp")
interv <- 24 #dosing interval, hours, only needed for glucuronidation models ("3CGlu", "Solve_3comp")
ndays <- 3 #number of days dosin is done, only needed for glucuronidation models ("3CGlu", "Solve_3comp")
ncores <- 4 #this is specific for the "3CGlu" to conduct parallelization
```

#load functions

```{r}
#All required R scripts and input files should be in the working directory
source("steadyState.R") # required for 1C model
source("glu_MaxConc.R") # required for "3CGlu"
source("CalcEAD.R") # required for 1C and 3CGlu
source("EADboxplot.R") # required for plotting
```

#load data

```{r}
chemical <- read.table(f0, sep = "\t", header = TRUE, quote = "", stringsAsFactors = FALSE) #Modify the input if needed for single quotes
chemical[1:2,]
```

```{r}
invitro <- read.table(f1, sep = "\t", header = TRUE)
invitro[1:2,]
```

#preparing data
Minor prep work is done on the data that comes from ICE 
Partitioning coefficients are only needed for the 3CGlu model as they are not calculated as part of the workflow here.
These can be obtained by running the "predict_partitioning_schmitt" function from the httk package.

```{r}
## input data assigned as chemical, column names should be labeled correctly 
## Following columns are required for glucuronidation models, 1C does not require partitioning coefficiets in order to run
## Funbound. should be fu, Kkidney2pu should be pkidney, Kliver2pu should be pliver, Krest2pu should be pbody
#colnames(chemical) <- gsub("Funbound.*", "fu", colnames(chemical))

#Converting units for internsic clearance values
chemical$Clint <- 10^chemical$Clint #this moves from log10 ul/ml/10^6 cells to just ul/ml/10^6 cells

#setting constants:
#dose = 1 #current calcualtions assume 1mg/kg/day

if (is.null(species)) {
  species <- "Human"
}
if (is.null(route)) {
  iv.dose <- TRUE
}
if (is.null(nsamples)) {
  nsamples <- 1000
}
if(!exists("dose") || is.null(dose)){
  dose <- 1
}
if(!exists("ConcentrationUnit") || is.null(ConcentrationUnit)){
  ConcentrationUnit <- "uM"
}
if(!exists("ncores") || is.null(ncores)){
  ncores<-detectCores() - 1
}
```

#if modelType given by the user is "1C"

```{r}
if (modelType == "1C") {    ### modelType needs to be selected as 1C in order for this step to work 
  
  chemInput <- chemical[,c("CASRN", "ChemicalName", "Clint", "fu", "MW")] ## subsetting chemical data for columns of interest
  
  CSS <- steadyState(inputData = chemInput, nsamples = nsamples, species = species, ConcentrationUnit = ConcentrationUnit) ## for this function to work SteadyState.R should be loaded in the enviornment
  
  EAD.out50 <- CalcEAD(Css = CSS[,c("CASRN", "50%", "fu")], inVitro = invitro, adj.fu = "fu") ## CalcEAD.R should be loaded in the enviornment
  colnames(EAD.out50) <- gsub("EAD","EAD.50", colnames(EAD.out50))
  EAD.out95 <- CalcEAD(Css = CSS[,c("CASRN", "95%", "fu")], inVitro = invitro, adj.fu = "fu")
  colnames(EAD.out95) <- gsub("EAD","EAD.95", colnames(EAD.out95))
  EAD.out <- join_all(list(EAD.out50,EAD.out95))
  EAD.out <- EAD.out[,setdiff(colnames(EAD.out), c("adj.fu", "fu", "adj.arm","arm"))] # remove the columns "fu", adj.fu","adj.arm" and "arm"
  
 #creating output file, should have EAD values and the parameters for calculating the circulating concentration
  
  CSS2 <- as.data.frame(CSS, stringsAsFactors = FALSE); colnames(CSS2) <- gsub("50%", "Css, 50%ile", colnames(CSS2)); colnames(CSS2) <- gsub("95%", "Css, 95%ile", colnames(CSS2))
  CSS2$Species<-species;CSS2$Model<-"1-compartment"; CSS2$"dose, mg/kg"<-dose; CSS2$"nSimulations" <- nsamples; CSS2$route<-route
 
   CSS2<-CSS2[,c("CASRN", "ChemicalName", "Css, 50%ile","Css, 95%ile", "Css_Unit", "Species","Model","dose, mg/kg", "route",
                "Clint", "fu", "MW")]
 
  ssEAD.out <- EAD.out[,c("CASRN", setdiff(colnames(EAD.out), c(colnames(invitro), "50%","95%")))]
  outputData <- join_all(list(CSS2,ssEAD.out))
  
  write.table(outputData, file = output_file, sep = '\t', col.names = TRUE, row.names = FALSE, quote = FALSE)#output file for user
  EADplot <- EADboxplot(EAD.out = outputData,label="EAD.50", species = species, route = route, modelType = modelType)
}
```


#if modelType given by the user is "3CGlu"
Note that this model requires some addtional physioloigical properties

```{r}
if (modelType == "3CGlu") { ## modelType needs to be selected as 3CGlu in order for this step to work 
  #Developing table for physical data
  physParam <- data.frame(species = c("human", "rat"), bw = c(70, 0.25), QC.mLmin = c(5600, 74), QliverC = c(0.26, 0.186), QkidneyC = c(0.22, 0.124), VplasmaC = c(0.043, 0.0312), VliverC = c(0.024, 0.078), VkidneyC = c(0.004, .0148), enterocytes = c(0.1223, 0.001407))
  #values for organ volumes from Davies et al 1993, table 2
  #these would ideally be supplied from user, defaults exist for the glucoronidated compound
  t1 <- data.frame(CASRN = chemical$CASRN, vmaxliver = NA, kmliver = NA,kuptakeC = NA, kGIingC = NA, kmgutg = 58400, vmaxgutgc = 361, fgutg = 1, fliverg = 1)
 #this would be data coming to ICE from OPERA; partitioning coeff could come from within
  chemParam <- as.data.frame(merge(chemical, t1), stringsAsFactors = FALSE)
  
  cl<-ncores; registerDoParallel(cl)
  Cmax.glu<- foreach(i = 1:nrow(chemParam),.combine = rbind) %dopar% {glu_MaxConc(physParam, chemParam[i,], Species = species, route =route, interv = interv , ndays = ndays, dose=dose, ConcentrationUnit = ConcentrationUnit)};
  
  #Calculating the EADs
  EAD.out_max <- CalcEAD(Css = Cmax.glu[,c("CASRN", "cmax")], inVitro = invitro)
   Cmax.glu$Species<-species;Cmax.glu$Model<-"3C Gluc"; Cmax.glu$Dose<-dose;
  
  ssEAD.out <- EAD.out_max[,c("CASRN", setdiff(colnames(EAD.out_max), c(colnames(invitro), "cmax")))]
  Cmax.glu$days <- ndays
  colnames(Cmax.glu)<-gsub("cmax", "Cmax", colnames(Cmax.glu)); colnames(Cmax.glu)<-gsub("Cmax_unit", "Cmax_Unit", colnames(Cmax.glu))
  Cmax.glu <- Cmax.glu[,c("CASRN", "ChemicalName", "Cmax", "Cmax_Unit", "Species", "Model", "dose, mg/kg", "route", "interv", "days")]
  
  outputData <- join_all(list(Cmax.glu,ssEAD.out))
  write.table(outputData, file = output_file, sep = '\t', col.names = TRUE, row.names = FALSE, quote = FALSE)#output file for user
  EADplot <- EADboxplot(EAD.out = outputData,label="EAD", species = species, route = route, modelType = modelType)
}
```

#if modelType given by the user is "Solve_3comp"

```{r}
if (modelType == "Solve_3comp") { ## modelType needs to be selected as Solve_3comp in order for this step to work 
  library(httk)
  #preprocessing variables:
  if (tolower(species) == "rat") {
    species_1 <- "Rat"
  }
  if (tolower(species) == "human") {
    species_1 <- "Human"
  }
  # Processing the chem information
  # ##will need to verify the labels on the opera data
  # chems <- as.data.frame(read.table(f0, header = TRUE, sep = "\t", stringsAsFactors = FALSE, quote ="",     na.strings = c("", "NA")))
  # inVitro <- read.table(f1, sep = "\t", header = TRUE)
  options(stringsAsFactors = FALSE)
  chem3cInput <- chemical[,setdiff(colnames(chemical), c("pkidney","pliver","pbody","Kgut2pu"))]#removing     the partitioning coefficience
  chem3cInput$logPwa <- 1*chem3cInput$HL #this gets the water octanal coeff
  #chemical$ClintNorm<-10^chemical$CLint #to get into correct units   #this is addressed in preprocessing

  #add chemical info to the table. Using variable coming from ICE

chem.physical_and_invitro.data <- add_chemtable(chem3cInput, current.table = chem.physical_and_invitro.data, data.list = list( CAS = "CASRN", Clint = "Clint", Funbound.plasma = "fu", pKa_Donor = 'pka_Donor', pKa_Accept = 'pka_Accept', logP = "LogP", logPwa = "logPwa", MW = "MW"), species = species_1, reference = paste0(species_1, "ICE"), overwrite = T)

  if (route == "oral") {
    iv.dose = FALSE
  } else {iv.dose = TRUE}
  
  EDcmax3day.3comp <- NULL
  cmax3dayivall <- NULL
  
  for (this.cas in chem3cInput[,'CASRN']) {
    outiv <- solve_3comp(chem.cas = this.cas, parameters = NULL, doses.per.day = 24/interv, days = ndays,tsteps = 4, daily.dose = 1, iv.dose = iv.dose, output.units = ConcentrationUnit, species = species_1, default.to.human = TRUE, plots = F, suppress.messages = TRUE)
    concMax <- max(outiv[,'Cplasma'])
    cmax3dayiv <- as.data.frame(cbind(this.cas,concMax, ConcentrationUnit))
    cmax3dayivall <- rbind(cmax3dayivall,cmax3dayiv)
    cmax3dayivall$concMax <- as.numeric(cmax3dayivall$concMax)
  }
  Cmax <- merge(chem3cInput, cmax3dayivall, by.x = "CASRN", by.y = "this.cas")
  #outcmax3dayiv <- ThreeC_httk(chemical, species = species_1, iv.dose)
 
  #Calculating the EADs
  EAD.out_max <- CalcEAD(Css = Cmax[,c("CASRN", "concMax")], inVitro = invitro)
  ssEAD.out <- EAD.out_max[,c("CASRN", setdiff(colnames(EAD.out_max), c(colnames(invitro), "Cmax")))]
  names(ssEAD.out) <- gsub("concMax", "Cmax", names(ssEAD.out) )
  
  names(Cmax) <- gsub("concMax", "Cmax", gsub("ConcentrationUnit", "Cmax_Unit", names(Cmax) ) )
  Cmax$route <- route
  Cmax$interv <- interv
  Cmax$days <- ndays
  Cmax$Species<-species; Cmax$Model<-"Solve_3comp"; Cmax$"dose, mg/kg"<-dose;
  Cmax <- Cmax[, c("CASRN", "ChemicalName", "Cmax", "Cmax_Unit", "Species", "Model", "dose, mg/kg","interv","days")]
  outputData <- join_all(list(Cmax,ssEAD.out))
  
  write.table(outputData, file = output_file, sep = '\t', col.names = TRUE, row.names = FALSE, quote = FALSE)#output file with results
  EADplot <- EADboxplot(EAD.out = outputData, label="EAD", species = species, route = route, modelType = modelType)
}
```


```{r}
sessionInfo()
```